---
title: "listArray"
author: 
- name: "Sigbert Klinke" 
  email: sigbert@hu-berlin.de
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{listArray}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, setup, echo=FALSE}
library(listArray)
```

# General

The aim of the package `listArray` is to create a data object which looks like an array, but behaves like a list.
Thus, something like 

* `x[letters[1:3]]` should address one element and
* at same time `x[letters[1:5]]` should be possible, too.

Additionally, any R object should possible as __index__, thus it should work:

* `x[1]`,
* `x["a"]`,
* `x[1,2]`,
* `x[c(1,2)]`,
* `x[mean]`, 
* `x[NULL]`, or
* `x[iris]`.

The package [hash](https://cran.r-project.org/package=hash) does something similar. But the keys follow
list naming convention and does not allow array indices.

Note that neither speed nor memory efficiency have any importance for the implementation.

# Generate a `listArray` from a vector, matrix or array

A `listArray` is a list (or environment), therefore list (or environment) operations can be applied:

```{r}
l <- listArray(1)
length(l)
names(l)
l[[1]]
class(l)
```

Thus, every R object which used in the index is translated to a unique name. The original indices can be obtained as string by:

```{r}
keys(l)
```

## Unnamed or named vectors

For creating vectors the `listArray` function can be used.

```{r}
# unnamed vectors
v <- 1:5
l <- listArray(v)
keys(l)
#
l <- listArray(letters[1:5])
l[1]
# named vector
v <- 1:5
names(v) <- letters[1:5]
l <- listArray(v)
l["a"]
```

## Matrix

For matrices or arrays:

```{r}
m <- matrix(1:9, 3, 3)
l <- listArray(m)
l[2,3]  # should be 8
```

## Names or numbers as indices

Since for `listArray`s `l[1]` and `l["A"]` is something different, you have to decide with named vectors, matrices
or arrays if you use the names or numbers. The default is to use names if available.

```{r}
m <- matrix(1:4, 2, 2)
colnames(m) <- LETTERS[1:2]
l <- listArray(m)
keys(l)
```

You can force with `use.names=FALSE` that always numerical indices will used

```{r}
m <- matrix(1:4, 2, 2)
colnames(m) <- LETTERS[1:2]
l <- listArray(m, use.names=FALSE)
keys(l)
```

## Ignore entries in a vector, matrix or array

Sometimes you may not want to store certain elements of vector, matrix or array; just think in terms of _sparse_ objects. 

```{r}
m <- diag(3)
l <- listArray(m, ignore=0)
keys(l)
```

The parameter `ignore` can be either a table of values to exclude or a function which returns for a vector a logical vector with `TRUE` (= value excluded) and `FALSE` (= value included).

```{r}
nozeroes <- function(v) { v==0 }
#
m <- diag(3)
l <- listArray(m, ignore=nozeroes)
keys(l)
```

## List or environment as base?

Rather than using a list it is possible to use an environment as base which might be of interest for package developers.

```{r}
e <- listArray(env=TRUE)
class(e)
e[1] <- "hello world"
e[1]
ls(e)
```
# Access a `listArray` object

## Extract operators

You simply use the `[` operator to access `listArray` elements.

```{r}
l <- listArray()
l[0] <- 1
l[0]
l[pi] <- pi
l[pi]
anotherpi <- pi
l[anotherpi]
l[1,-2] <- 3
l[1,-2]
```

## Difference between `listArray` and vector

A `listArray` considers each index element as different. The following works for vectors:

```{r}
m <- 1:5
m[1:2]
```

But `l[1:2]` returns `NULL` since the index `1:2` does not exist.

```{r}
l <- listArray(m)
l[1:2]
keys(l)
```


## Difference between `listArray` and matrix/array

Similarly, it holds

```{r, error=TRUE}
m <- matrix(1:4, 2, 2)
m[1,]
l <- listArray(m)
l[1,] # will even throw an error
```

## Normalization and the options `listArray.XXX`

To achieve, e.g. that `l[1:3]` and `l[c(1,2,3)]` address the same element, as we would expect, we need some kind of normalization. Since `1:3` and  `c(1,2,3)` a different R objects, a normalization is internally done.

```{r}
identical(1:2, c(1,2)) # delivers FALSE!
# but
m <- matrix(1:9, 3, 3)
m[1:2,2]
m[c(1,2),2]
```

There are two problems

1. `1:3` is of class `integer` whereas `c(1,2,3)` is of class `numeric` and
2. `1:3` is a _compact_ sequence in R whereas `c(1,2,3)` is a _full_ vector

Therefore, normalization currently consists of 

1. converting in indices everything from class `integer` to `numeric` and
2. expand compact sequences like `1:3` to real vectors, e.g. `c(1,2,3)`.

The normalization steps can be switched off by setting the options `listArray.expand` and `listArray.int2num`.

```{r}
l <- listArray()
l[1:3] <- 1
l[c(1,2,3)]
options(listArray.expand=FALSE) # now 1:3 != c(1,2,3)
l <- listArray()
l[1:3] <- 1
l[c(1,2,3)]
```

The default is that `listArray.expand` and `listArray.int2num` are not set which is interpreted as `listArray.expand=TRUE` and `listArray.int2num=TRUE`.

# The workhorse function `key`

The main function to create a string from a set of R objects is `key`. By  using `l[...]` internally
 is called `l[[key(...)]]`. Thus, you could only use `key` rather than a `listArray` object.

The normalization are done via

1. `rapply(l, expand, classes=c("numeric", "integer"), how="replace")` with `expand  <- function(x) { unserialize(serialize(x, connection=NULL, version=2)) }` and 
2. `rapply(l, as.numeric, classes="integer", how="replace")`.

In future might be further normalization necessary then the two above. 

# Acknowledgments

Thanks to [Henrik Bengtsson](https://www.mail-archive.com/r-help@r-project.org/msg259696.html) and [Duncan Murdoch](https://www.mail-archive.com/r-help@r-project.org/msg259700.html) which hinted me how to normalize a compact sequence in R without writing C++ code.




